# Sampler Application Example 2 - Single threshold parameter

In this second example, we will use a single `b` threshold parameter (the evidence required to make a response).
Altering the code/model is simple. First, update your parameter object (e.g.`pars`) by including a single `b` threshold parameter:
```{r eg2pars}
pars <- c("b", "A", "v1", "v2", "t0")
```

The `priors` object will update based on the length of your parameter object. <b>Remove this?</b>
```{r eg2priors}
priors <- list(theta_mu = rep(0, length(pars)),
  theta_sig = diag(rep(1, length(pars)))
)
```
The next step is to modify your log likelihood function by updating the threshold parameter vector (`bs`) so that the name and number of threshold parameters (`b`) matches the name and number of threshold parameters you've specified in your `pars` vector. In this case, we require a single `b` parameter.
``` {r bslbaLL, echo=TRUE, eval=FALSE}
  bs <- x["A"] + x["b"][data$condition]
```
The subsequent lines of code in the log likelihood function remain unchanged.

As mentioned in example 1, <b>ADD A CROSS REFERENCE HERE</b> you may set model start points. It is important to provide the same number of values for `mu` as the number of parameters you've set in your `pars` vector.
```{r eg2modStartPoints, echo=TRUE, eval=FALSE}
start_points <- list(
  mu = c(.2,.4, .3, 1.3, -2), # We have set five parameter start points here, 
                              # matching the number of parameters in the pars vector.
  sig2 = diag(rep(.01, length(pars)))
)

```
And now you're ready to run the sampler as outlined in example 1<b>ADD A CROSS REFERENCE HERE</b>!
