# PMwG sampler and sequential sampling models {#forstmannChapter}


In this chapter we'll demonstrate how to use the PMwG sampler with a sequential sampling model; the Linear Ballistic Accumulator (LBA).
Please ensure the `PMwG` and `rtdists` packages are installed.
```{r loadpkg}
library(pmwg)
library(rtdists)
```

## The speed-accuracy tradeoff in perceptual decisions

We demonstrate the application of the LBA with the PMwG sampler to a study of perceptual decision making (Forstmann et al., 2008). Forstmann et al looked at neural correlates of decision making under time pressure, with an aim to identify areas of the brain associated with speed-accuracy tradeoff. Imaging (fMRI) and behavioural data was collected; however, we will analyse behavioural data from the decision-making task only. In terms of modelling the data, Forstmann expected to find differences in thresholds for each of the three speed-emphasis conditions. We have included the Forstmann et als data in the `PMwG` package as a data frame named `forstmann`. The sampler requires a data frame with a <b>`subject`</b> column. The subject column data type can be a factor or numeric.

Table \@ref(tab:forsthead10) shows the first ten trials from the Forstmann dataset. Participants `(n = 19)` were asked to indicate whether a cloud of dots in a random-dot kinematogram (RDK) moved to the left or the right of the screen. The IV was a within-subject, speed-accuracy manipulation where, before each trial began, pariticipants were instructed to make their choice <i>accurately</i> `(condition = 1)`, with <i>urgency</i>`(condition = 3)`or were presented with a <i>neutral</i> message `(condition = 2)`. Stimuli moved either <i>left</i> `(stim = 1)` or <i>right</i> `(stim = 2)` and responses were <i>left</i> `(resp = 1)` or <i>right</i> `(resp = 2)`. Response times `(rt)` were recorded in seconds. For more information about the design of the experiment please see [the original paper](https://www.pnas.org/content/105/45/17538.short).

```{r forsthead10, echo=FALSE, out.width='80%', fig.asp=.75, fig.align='center', warning=FALSE}
kable(head(forstmann, 10), row.names = F,  align = "ccccc", caption = 'First 10 trials in Forstmann dataset. The `forstmann` dataset is an data frame ')
```

## Linear Ballistic Accumulator Parameters

There are preliminary steps we need to complete before running the sampler. Let's begin by defining the Linear Ballistic Accumulator (LBA) [@brown2008simplest] model parameters.

* `b` threshold parameter (the evidence required to trigger a response)
* `v` is the mean drift rate or average speed of evidence accumulation
* `A` is the range of start points for accumulators 
* `t0` is non-decision time 
* `sv` is the standard deviation of the across-trial distribution of drift rates

## Writing the log-likelihood function {#writellFunc}

Just as we did with the SDT example, we'll write a slow and a fast log-likelihood function. The runtime difference is caused by calling the `dLBA` function line-by-line for the slow log-likelihood and calling the `dLBA` function once for all the data in the fast log-likelihood function. When writing a new log-likelihood function, we suggest starting with a slow, line-by-line function for easier debugging. 


The LBA log-likelihood function takes three arguments: 

* `x` is a named parameter vector (e.g. `pars`)
* `data` is your dataset (e.g.`forstmann`). Your dataset must include a `"subject"` column
* `sample = FALSE` calculates a density function or `TRUE` generates a posterior predictive sample that matches the shape of data.


The log-likelihood function shown below includes functions from the `rtdists` package for generating data and estimating density. If you'd like to run through this example, it is best to copy the `tw_lba_ll` function from the code block below rather than copying from the separate code chunks where curly braces have been removed.

NOTE: The trialwise log-likelihood is very slow and inneficient because `rLBA` and `dLBA` will be called on each line of the data. This will result in very slow sampling times and is a consequence of the `rtdists` package, not an issue with the PMwG sampling speed. If you have experience writing log-likelihoods, we recommend writing a faster version than our trialwise function, or use the fast log-likelihood we have written in section \@ref(fstLBALL). If you are new to modelling, we recommend trying the trialwise (slow) log-likelihood function as it is easier to follow, troubleshoot and is less likely to result in errors. 

Let's begin by loading the `rtdists` package...

```{r instrtdist}
library(rtdists)
```

and now our complete trialwise (slow) log-likelihood function.

```{r trialwiseLL, attr.source = '.numberLines', include=TRUE }
tw_lba_ll <- function(x, data, sample = FALSE) {
  x <- exp(x)
  if (any(data$rt < x["t0"])) {
    return(-1e10)
  }
  
  if (sample) {
    tmp <- numeric(nrow(data))
    data$rt <- rep(NA, nrow(data))
    data$resp <- rep(NA, nrow(data))
  } else {
    out <- numeric(nrow(data))
  }
  
  for (i in 1:nrow(data)) {
    A = x["A"]
    b = x[paste0("b.", data$condition[i])] + A
    vc = x["vc"]
    ve = x["ve"]
    t0 = x["t0"]
    s = c(1, 1)
    
    if (data$stim[i] == 1) {
      vs = c(vc, ve)
      } else {
      vs = c(ve, vc)
      }
    
    if (sample) {
      tmp <- rLBA(n = 1,
                  A = A,
                  b = b,
                  mean_v = vs,
                  sd_v = s,
                  t0 = t0,
                  dist = "norm",
                  silent = TRUE
                  )
      data$rt[i] <- tmp$rt
      data$resp[i] <- tmp$resp
    } else {
      out[i] <- dLBA(rt = data$rt[i],
                     response = data$resp[i],
                     A = A,
                     b = b,
                     mean_v = vs,
                     sd_v = s,
                     t0 = t0,
                     dist = "norm",
                     silent = TRUE
      )
      }
  }
  
if (sample) {
    return(data)
  } else {
    bad <- (out < 1e-10) | (!is.finite(out))
    out[bad] <- 1e-10
    out <- sum(log(out))
    return(out)
  }
}
```


The first line in the `tw_lba_ll` function (Line 2 below) takes the exponent of the parameter values. We do this as the LBA requires positive parameter values that are on the real line. Line 3 and 4 then checks RTs are faster than the non-decision time parameter `t0`, and returns a low value if `t0` is larger than RT, indicating that the given value of `t0` is unlikely. 

```{r twLBAll1, attr.source = '.numberLines startFrom="2"', echo=TRUE, eval=FALSE}
  x <- exp(x)
  if (any(data$rt < x["t0"])) {
    return(-1e10)
```


Now we create a vector with values sampled from the posterior distribution OR estimating the density. If `sample = TRUE`, we remove all responses `(resp)` and rts.  This means when we return `data`, we are returning the posterior predictive data which matches with the associated `subject` and `condition` and then generate them from the random function of the model.

If `sample = FALSE` (the `else` statement from line 11) we create an `out` vector, with a length equal to the number of rows in the dataset, and store the likelihood value for each subject and condition.

```{r twLBAll2, attr.source = '.numberLines startFrom="7"', echo=TRUE, eval=FALSE}
  if (sample) {
    tmp <- numeric(nrow(data))
    data$rt <- rep(NA, nrow(data))
    data$resp <- rep(NA, nrow(data))
  } else {
    out <- numeric(nrow(data))
  }
```


Next, we loop over rows in the dataset. In this `for` loop, we find the values `(x)` of each parameter in our model for each row, so that any conditional parameters (for example `b` in our model) are correctly assigned. For example, we want to calculate the density for a model that has three threshold parameters (one for each condition; `1  = accuracy`, `2 = neutral`, or `3 = speed`). In the loop, we paste `b.` to the condition in row `[i]` and add `A` (the start point - we do this to ensure the threshold is greater than the starting point). 

On line 22 we set the order of our drift rate parameters. Recall that `stim = 1` is a stimulus moving to the left. `dLBA` requires the drift accumulators to be matched i.e. when `data$stim[i] == 1`, the drift rate for the correct accumulator `(vc)` is in position one, so we order the drift rates; `vs = c(vc, ve)`. The `else` statement addresses right moving stimuli `data$stim[i] == 2`, the incorrect accumulator `(ve)` is the first accumulator, so the drift rate parameter order is `vs = c(ve, vc)`. This ensures that the correct `(vc)` and error `(ve)` drift rates match with the corresponding accumulators for given stimuli. 

```{r twLBAll3, attr.source = '.numberLines startFrom="15"', echo=TRUE, eval=FALSE}
  for (i in 1:nrow(data)) {
    A = x["A"]
    b = x[paste0("b.", data$condition[i])] + A
    vc = x["vc"]
    ve = x["ve"]
    t0 = x["t0"]
    s = c(1, 1)
    if (data$stim[i] == 1) {
      vs = c(vc, ve)
    } else {
      vs = c(ve, vc)
    }
```

The following section calls the relevant `rtdists` function depending on whether we are sampling from the posterior predictive distribution `(rLBA)` or estimating the density `(dLBA)`. We then input the parameters from above (using the names set above) into the relevant function. When generating data from the posterior predictive distribution (Line 30-37), `rLBA` is called for each line of the data, storing the generated `rt` and `response` given the posterior parameter estimates in the `tmp` vector (which we then reassign to the empty `data$rt` and `data$resp` columns). We set `n = 1`, since we are calling `rLBA` on 1 row of the data. When estimating the density (Line 42 to 50), dLBA is called for each line of the data, storing the probability of the `rt` and `response` under the proposed parameters `(x)` in the `out` vector. 


```{r lbaLL3, attr.source='.numberLines startFrom="29"', echo=TRUE, eval=FALSE}
  if (sample) {
      tmp <- rLBA(n = 1,
                  A = A,
                  b = b,
                  mean_v = vs,
                  sd_v = s,
                  t0 = t0,
                  dist = "norm",
                  silent = TRUE
                  )
      data$rt[i] <- tmp$rt
      data$resp[i] <- tmp$resp
    } else {
      out[i] <- dLBA(rt = data$rt[i],
                     response = data$resp[i],
                     A = A,
                     b = b,
                     mean_v = vs,
                     sd_v = s,
                     t0 = t0,
                     dist = "norm",
                     silent = TRUE
      )
      }
```

This final section tells the function what to return; `data` - when sampling posterior predictive data (Line 56) - or the sum of the likelihoods - when estimating density (Line 57 - 61). On line 58 we take all implausible likelihood values, assign them to the `bad` object and then (line 59) set them to an extremely unlikely value, to prevent numerical errors. The final two lines within the `else` statement take the log of all likelihood values, sums them, assigns the model’s log-likelihood value to the `out` variable and returns that value.

```{r lbaLL43, attr.source='.numberLines startFrom="55"', echo=TRUE, eval=FALSE}
if (sample) {
    return(data)
  } else {
    bad <- (out < 1e-10) | (!is.finite(out))
    out[bad] <- 1e-10
    out <- sum(log(out))
    return(out)
  }
```

### Fast LBA Log-likelihood Function {#fstLBALL}

As the data is large, and the `dLBA` function takes some time to run, the log-likelihood code above is computationally inefficient. There are several ways to improve the log-likelihood's performance; in our example below, we reduce the number of calls to `dLBA` to one call. We do this by passing a list of `dLBA` parameter values for the length of the data. 

<b>Note:</b> When generating posterior predictive data, the `rLBA` function is still executed for each row of data; however, it is only executed several times, so computational efficiency is uncompromised.


```{r fast_lba_ll, include=TRUE}
fast_lba_ll3b <- function(x, data, sample = FALSE) {
  x <- exp(x)
  if (any(data$rt < x["t0"])) {
    return(-1e10)
  }
  
  if (sample) {
    data$rt <- rep(NA, nrow(data))
    data$resp <- rep(NA, nrow(data))
  } else {
    out <- numeric(nrow(data))
  }
  
  if (sample) {
    for (i in 1:nrow(data)) {
      A = x["A"]
      b = x[paste0("b.", data$condition[i])] + A
      vc = x["vc"]
      ve = x["ve"]
      t0 = x["t0"]
      s = c(1, 1)
      
      if (data$stim[i] == 1) {
        vs = c(vc, ve)
      } else {
        vs = c(ve, vc)
      }
      
      tmp <- rLBA(n = 1,
                  A = A,
                  b = b,
                  mean_v = vs,
                  sd_v = s,
                  t0 = t0,
                  dist = "norm",
                  silent = TRUE
                  )
      data$rt[i] <- tmp$rt
      data$resp[i] <- tmp$resp
    }
  } else {
    all_b <- numeric(nrow(data))
    vlist <- list("v.1" = numeric(nrow(data)), 
                 "v.2" = numeric(nrow(data)))
    stim <- levels(data$stim)
    con <- levels(data$condition)
    
    for (c in con) {
      for (s in stim) {
        use <- data$condition == c & data$stim == s
        if (any(use)) {
          bs = x[paste0("b.", c)] + x["A"]
          all_b[use] = bs
          vc = x["vc"]
          ve = x["ve"]
          if (s == 1) {
            vlist$v.1[use] = vc
            vlist$v.2[use] = ve
          } else {
            vlist$v.1[use] = ve
            vlist$v.2[use] = vc
          }
        }
      }
    }
    
    out <- dLBA(rt = data$rt,
                response = data$resp,
                A = x["A"],
                b =  all_b,
                mean_v = vlist,
                sd_v = c(1, 1),
                t0 = x["t0"],
                distribution = "norm",
                silent = TRUE
    )
    }
  
  if (sample) {
    return(data)
  } else {
    bad <- (out < 1e-10) | (!is.finite(out))
    out[bad] <- 1e-10
    out <- sum(log(out))
    return(out)
  }
}
```

You should improve your log-likelihood's performance as you see fit.  When you're confident that your log-likelihood code functions correctly, we suggest saving it as a separate script so it can be sourced and loaded when running the sampler. If you're learning how to write log-likelihood functions, take a look at our [troubleshooting section](#troubleshoot) for tips. In our experience, a very large proportion of problems with sampling and inference are caused by inadequate checking and care in the likelihood function.

## PMwG Framework

Now that we have a log-likelihood function, we can set up the `PMwG` sampler. Running the sampler follows the same procedure outlined in the [SDT chapter](#sdtPMwG); we need to set up a vector of model parameter names, create a `priors` object, source our LBA log-likelihood script and then create our `sampler` object.


Let's begin by creating a vector of model parameter names, which we'll use in our log-likelihood function. You can name this object as you wish; however, in our example, we name it `pars`. ^[The parameters you list in the `pars` vector <b>must</b> match those included in your log-likelihood function i.e. they must have the same names and the same number of parameters. Refer to the [troubleshooting section](#troubleshoot) for more detail.] For the `forstmann` dataset, we use three threshold parameters (one `b` for each condition) because we assume that the condition has an effect on level of caution, we include two drift rate parameters: `ve` for the incorrect accumulator and `vc` for the correct accumulator, a start point parameter `A` and a non-decision time `t0` parameter. 

We've made a decision to set the `sv` to 1 to satisfy the scaling properties of the model. As such, we haven't included the `sv` parameter in the `pars` vector - it is found in the LBA's log-likelihood function (see above).

```{r pars}
pars <- c("b1", "b2", "b3", "A", "ve", "vc", "t0")
```


For the mean of the distribution for random effects (theta_mu), we assume a multivariate normal prior. The user can specify the mean and variance of this prior distribution using the object `priors`, which has elements `theta_mu_mean` and `theta_mu_var`. A typical setting for LBA models is to set `theta_mu_mean` to be a zero vector and to set `theta_mu_var` to be a multiple of the identity matrix, e.g. with 9 on the diagonal (representing a standard deviation of 3 for the subject-level means in the prior). 

We create our `priors` object; a list that contains two components:

* `theta_mu_mean` a vector containing the prior for model parameter means
* `theta_mu_var` the prior covariance matrix for model parameters.


```{r priors}
priors <- list(theta_mu_mean = rep(0, length(pars)),
  theat_mu_var = diag(rep(1, length(pars)))
)
```

Now source and load your log-likelihood script <i>before</i> you create the sampler object.
```{r loadLL, echo=TRUE, eval=FALSE}
source(file = "fast_lba_ll3b.R")
```

Next we specify the PMwG `sampler` object. 
The `pmwgs` function takes a set of arguments (listed below) and returns a list containing the required components for performing the particle metropolis within Gibbs steps.

* `data =` your data - a data frame (e.g.`forstmann`) with a column for participants called <b>`subject`</b>
* `pars =` the model parameters to be used (e.g.`pars`)
* `prior =` the priors to be used (e.g.`priors`)
* `ll_func =` name of log-likelihood function to be used (e.g.`fast_lba_ll`)

```{r forstSampObj, eval=FALSE}
sampler <- pmwgs(data = data,
                 pars = pars,
                 prior = priors,
                 ll_func = fast_lba_ll3b
                 )
```


### Model start points {#lbaStartPts}
There is also an option to set model start points. We have specified sensible start points for the `forstmann` dataset under the LBA model. If you choose not to specify start points, the sampler will randomly sample points from the prior distribution. 

The `start_points` object contains two vectors:

* `mu` a vector of start points for the mu of each model parameter 
* `sig2` vector containing the start points of the covariance matrix of covariance between model parameters.

<b>Note:</b> <i>Start points must be on the real line. Our log-likelihood function immediately takes the exponent of the start points and only returns positive values, so we use the log of sensible start points here.</i> 

```{r lbaStartPts, echo=TRUE, eval=FALSE}
start_points <- list(mu = log(c(1.2, 1.2, 1.2,
                                1.4, 1.3, 3.5, 
                                0.13)
                              ),
                     sig2 = diag(rep(.01, length(pars)))
                     )
```


### Running the sampler {#run-sampler}

Setup is now complete and we can run the sampler. First, we use the `init` function to generate initial start points for the random effects and store them in the `sampler` object. Here we specify start points by providing values for the `start_mu` and `start_sig` arguments.

```{r runlbaSampler, echo=TRUE, eval=FALSE}
sampler <- init(sampler, 
                start_mu = start_points$mu,
                start_sig = start_points$sig2
                )
```

<br>
To run the sampler, we use the `run_stage` function. To execute the `run_stage` function, you <b>must</b> provide values for two arguments: 


* `pmwgs =` the `sampler` object including parameters that were created by the `init` function above.
* `stage =` the sampling stage (In order; `"burn"`, `"adapt"` or `"sample"`).

The following arguments listed below are optional:


* `iter =` is the number of iterations for the sampling stage. For burn-in, it is important to have enough iterations that the chains converge on the posterior distribution. Default = 1000.
* `particles =` is the number of proposals (particles) generated for each random effect, on each iteration. Default = 1000, but set smaller or larger in order to target a reasonable acceptance rate (i.e. 10-60%).
* `display_progress =` display a progress bar during sampling
* `epsilon =` is a value greater than 0 which scales the variance of the proposal distribution. Smaller values (i.e. narrower proposal distributions) can lead to higher acceptance rates, but slower coverage of the posterior. Smaller values are especially useful when the number of random effects is large (e.g. >10). Default = 1.
* `n_cores =` the number of cores on a machine you wish to use to run the sampler. This allows sampling to be run across cores (parallelising for subjects). Default = 1. <b> Note:</b> Setting `n_cores` greater than 1 is only permitted on Linux and Mac OS X machines.


<br>
The first sampling stage is burn-in `"burn"`. The burn-in stage allows time for the sampler to move from the (arbitrary) start points that were provided by the user to the mode of the posterior distribution. This can be checked by examining the chains for stationarity. We take the `sampler` object created in the `init` function above, set the stage argument to `"burn"` and assign the outcome to an object called `burned`. 

```{r forstlbaburn, echo=TRUE, eval=FALSE}
burned <- run_stage(sampler, 
                    stage = "burn",
                    iter = 500,
                    particles = 2000,
                    epsilon = .5
                    )
```


Next is the adaptation stage `"adapt"`. The adaptation stage draws samples using a simple, but relatively inefficient proposal distribution (the same proposal distribution as the `"burn"`stage). Enough samples are drawn to allow the algorithm to estimate a much more sophisticated and efficient proposal distribution, using conditional normal distributions. We take the `burned` object created in the previous stage and set iterations `iter =` to a high number (e.g. `10000`), as it should exit before reaching this point. If it doesn't, there is likely an issue with acceptance rates, the likelihood function or limited data to operate on (i.e. few trials in some conditions). Here, we have saved the outcome of the adaptation stage to an object called `adapted`.

```{r adaptation, echo=TRUE, eval=FALSE}
adapted <- run_stage(burned, 
                     stage = "adapt", 
                     iter = 10000, 
                     particles = 2000,
                     epsilon = .5
                     )
```

The final stage is the sampling stage `"sample"`. The sampling stage uses the sophisticated and adaptive conditional normal proposal distributions. This allows for very efficient sampling, using far fewer particles. Samples from this stage are taken from the 'posterior distribution' and stored in the `sampled` object. 

``` {r part 7.12,eval=FALSE}
sampled <- run_stage(adapted, 
                     stage = "sample",
                     iter = 1000, 
                     particles = 200,
                     epsilon = .5
                     )
```


The `sampled` object includes all samples from the `"sample"` stage above and the following elements:

* `data` : your data (data frame) you included in your analysis
* `par_names`: parameter names 
* `n_pars`: number of parameters
* `n_subjects`: number of subjects
* `subjects`: subject IDs (1:n)
* `prior`: list of the prior used
* `ll_func`: the likelihood function specified
* `samples`:
* `alpha`: three dimensional array of random effects draws (dim = parameters x subjects x samples)
* `theta_mu`: two dimensional array of parameter draws (dim = parameters x samples)
* `theta_sig`: three dimensional array of covariance matrix draws (dim = covariance x samples)
* `stage`: specifies the stage the sample is from (length = samples)
* `subj_ll`: likelihood value for each subject for each iteration (dim = subject x samples)
* `a_half`: the parameter used in calculating the inverse Wishart (dim = parameters x samples)
* `idx`: total number of samples
* `last_theta_sig_inv`: the inverse of the last sample for theta_sig (the variance-covariance matrix).


You should save your sampled object at this point.
```{r}
save(sampled, file = "forst3bSamp.RData")
```

## Simulating Posterior Predictive Data {#genppdatafunc}

We can generate posterior predictive data by setting `sample = TRUE` in our log-likelihood function to generate response times and responses <i>given</i> the posterior parameter estimates for each subject. To do this, we use the `gen_pp_data` function below, which calls our log-likelihood function embedded in our `sampled` object. The `gen_pp_data` function takes four arguments:

* `sampled`: is the object/output from the PMwG sampler
* `n`: the number of posterior samples
* `ll_func =`: the log-likelihood function embedded in the `sampled` object
* `rbind.data = `: bind the rows of each predictive sample into a rectangular array, or leave as a list.

```{r genPPDataFunc, Include=TRUE}
gen_pp_data <- function (sampled, n, ll_func = sampled$ll_func, rbind.data = TRUE) {
  sampled_stage <- length(sampled$samples$stage[sampled$samples$stage == "sample"])
  iterations <- round(seq(from = (sampled$samples$idx - sampled_stage),
                            to = sampled$samples$idx,
                            length.out = n))
  data <- sampled$data
  S <- sampled$n_subjects
  pp_data <- list()
  
    for (s in 1:S){
       print(paste0("subject", s))
      for (i in 1:length(iterations)) {
        print(i)
        x <- sampled$samples$alpha[, s, iterations[i]]
        names(x) <- pars
        out <- ll_func(x = x, 
                       data = data[data$subject == s, ],
                       sample = TRUE)
        
        if (i == 1){
          pp_data[[s]] = cbind(pp_subj = i, out)
        }
        else {
          pp_data[[s]] = rbind(pp_data[[s]],
                         cbind(pp_subj = i, out))
        }
      }
    }
  
    if (rbind.data){
    tidy_pp_data <- do.call(rbind, pp_data)  
    return(tidy_pp_data)
    }
    else {
      return(pp_data)
    }
}
```

```{r loadForstSampled, include = FALSE}
load("forstmann3b_sampled.RData")
load("ppdata3b.RData")
```

We generate 20 posterior predictive data samples. 
```{r genppdata, eval=FALSE}
pp_data_3b <- gen_pp_data(sampled, n = 20)
```

The returned data is a matrix with the same dimensions and names as `forstmann` -- with the addition of `pp_iter` column. `pp_iter` is the iteration of posterior sample (in this example `i = 1:20`) for the corresponding subject. We now have two matrices based on samples from either model. The response (`resp`) and response time (`rt`) columns now contain posterior predictive data.

In the next section, we will use the posterior predictive data to assess descriptive adequacy.

### Assessing Descriptive Adequacy (goodness of fit)

Now we will plot the posterior predictive data against the real data. In the section below we compare observed RTs against predicted RTs, which is common for RT modelling; however, the code could also be modified for different types of data.

```{r plotppdata3b, include=TRUE}
# Subject x condition Q25, median and Q75 respone time + mean accuracy 
# Forstmann dataset
pq3b <- forstmann %>% 
  group_by(condition, subject) %>% 
  summarise(Q25 = quantile(rt, prob = 0.25),
            median = median(rt), 
            Q75 = quantile(rt, prob = 0.75),
            acc =  mean(ifelse(stim == resp, 1, 0)),
            .groups = "keep"
            )

# Subject x condition Q25, median and Q75 respone time for posterior predictive data
pp_pq3b <- pp_data_3b %>%
  group_by(condition, pp_iter, subject) %>%
  summarise(Q25 = quantile(rt, prob = 0.25),
            median = median(rt), 
            Q75 = quantile(rt, prob = 0.75),
            acc = mean(ifelse(stim == resp, 1, 0)),
            .groups = "keep"
            )

# Combine data with posterior predictive data and add data source
pq3b <- bind_rows(cbind(src = rep("data", nrow(pq3b)), pq3b),
                cbind(src = rep("model", nrow(pp_pq3b)), pp_pq3b)
                )

# Mean Q25, median, Q4 and accuracy for data and posterior predictive data
av_pq3b <- pq3b %>%
  group_by(src, condition) %>%
  summarise_at(vars(Q25:acc), mean)

# Variances of posterior samples
pp_var3b <- pq3b %>%
  filter(src != "data") %>%
  group_by(condition, pp_iter, src) %>%
  summarise_at(vars(Q25:acc), mean)

# Convert source column to a factor and add labels
av_pq3b$src <- factor(av_pq3b$src,
                    levels = c("model", "data"),
                    labels = c("model", "data")
                    )

pp_varsrc3b <- factor(pp_var3b$src,
                    levels = c("model", "data"),
                    labels = c("model", "data")
                    )
# Rename conditions
levels(av_pq3b$condition) <-
  levels(pp_var3b$condition) <- c("Accuracy", "Neutral", "Speed")

# Convert rt to milliseconds and acc to percentage
av_pq3b$acc <- 100 * av_pq3b$acc
pp_var3b$acc <- 100 * pp_var3b$acc
av_pq3b[, c("Q25", "median", "Q75")] <- 1000 * av_pq3b[, c("Q25", "median", "Q75")]
pp_var3b[, c("Q25", "median", "Q75")] <- 1000 * pp_var3b[, c("Q25", "median", "Q75")]
```

```{r hideppdataplots, echo=FALSE}
# Plot response times for data and posterior predictive data
rtplot3b <- ggplot(data = av_pq3b,
                 mapping = aes(x = condition,
                               y = Q25,
                               group = src,
                                colour = src
                               )
                 ) +
  geom_point(size = 2) +
  geom_point(mapping = aes(x = condition,
                           y = median,
                           color = src,
                           group = src
                           ),
             size = 4) +
  geom_point(mapping = aes(x = condition,
                           y = Q75,
                           color = src,
                           group = src
                           ),
             size = 2) +
  geom_point(data = pp_var3b,
             mapping = aes(x = condition,
                           y = Q25,
                           group = src),
             size = 1,
             color = "black",
             alpha = 0.05) +
  geom_point(data = pp_var3b,
             mapping = aes(x = condition,
                           y = median,
                           group = src),
             size = 2,
             color = "black",
             alpha = 0.05) +
  geom_point(data = pp_var3b,
             mapping = aes(x = condition,
                           y = Q75,
                           group = src),
             size = 1,
             color = "black",
             alpha = 0.05) +
  labs(x = "Condition",
       y = "RT Percentiles (ms)",
       colour = "Data source") +
  theme(axis.title.y = element_text(size = 14, face = 'bold'),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.75, 'cm'),
        legend.title = element_text(size = 14, face = 'bold'),
        legend.title.align = 0.5,
        legend.position = "top") +
   ylim(275, 650)

# Plot accuracy for data and posterior predictive data
accplot3b <- ggplot(data = av_pq3b,
                  mapping = aes(x = condition,
                                y = acc,
                                color = src,
                                group = src
                                )) +
  geom_point(size = 4) +
  geom_point(data = pp_var3b,
             mapping = aes(x = condition,
                           y = acc,
                           group = src),
             color = "black",
             alpha = .2,
             size = 2) +
  labs(x = "Condition",
       y = "Mean Accuracy (%)") +
  theme(axis.title.y = element_text(size = 14, face = 'bold'),
        axis.text.y = element_text(size=14),
        axis.title.x = element_text(size = 14, face = 'bold'),
        axis.text.x = element_text(size=14),
        legend.position = "none") +
  ylim(75, 90)
```  

```{r ppdataplotarrg3b, fig.align='center', fig.height=10, echo=FALSE, warning=FALSE}
plots3b <- rtplot3b/accplot3b
plots3b
```

##Evaluating different models - single threshold LBA 

Now let's run through an example where the model differs and we assume that participants's level of caution does not change with emphasis instructions i.e. the threshold for `Accuracy`, `Neutral`, and `Speed` conditions is the same. We will call this our "single threshold LBA model". The single threhold LBA model's log-likelihood has one `b` parameter, instead of three (see line 18, 31 and 64 below).

For brevity, we specify only the '<i>fast</i>' log-likelihood function for the single threshold model. 
```{r 1bfastll, attr.source = '.numberLines', include=TRUE}
fast_lba_ll1b <- function(x, data, sample = FALSE) {
  x <- exp(x)
  if (any(data$rt < x["t0"])) {
    return(-1e10)
  }
  
  if (sample) {
    tmp <- numeric(nrow(data))
    data$rt <- rep(NA, nrow(data))
    data$resp <- rep(NA, nrow(data))
  } else {
    out <- numeric(nrow(data))
  }
  
  if (sample) { 
    for (i in 1:nrow(data)) {
      A = x["A"]
      b = x["b"] + A
      vc = x["vc"]
      ve = x["ve"]
      t0 = x["t0"]
      s = c(1, 1)
      if (data$stim[i] == 1) {
        vs = c(vc, ve)
      } else {
        vs = c(ve, vc)
      }
      
      tmp <- rLBA(n = 1,
                  A = A,
                  b = b,
                  mean_v = vs,
                  sd_v = s,
                  t0 = t0,
                  dist = "norm",
                  silent = TRUE
                  )
      data$rt[i] <- tmp$rt
      data$resp[i] <- tmp$resp
    } 
  } else {

  vlist = list("v.1" = numeric(nrow(data)), 
               "v.2" = numeric(nrow(data)))
  
  
  for (c in levels(data$condition)) {
    for (s in levels(data$stim)) {
      use <- data$condition == c & data$stim == s
        vc = x["vc"]
        ve = x["ve"]
        if (s == 1) {
          vlist$v.1[use] = vc
          vlist$v.2[use] = ve
        } else {
          vlist$v.1[use] = ve
          vlist$v.2[use] = vc
        }
    }
  }
    out <- dLBA(rt = data$rt,
                response = data$resp,
                A = x["A"],
                b = x["b"] + x["A"],
                mean_v = vlist,
                sd_v = c(1, 1),
                t0 = x["t0"],
                distribution = "norm",
                silent = TRUE)
    }
  
  if (sample) {
    return(data)
  } else {
    out<-sum(log(pmax(out, 1e-10)))
    return(out)
  }
}
```

### PMwG framework for a single threshold model
The PMwG sampler procedure remains the same for all models. For our three threshold LBA model, we need the updated log-likelihood function from above, an updated parameter vector, start points and priors.

```{r part9.8,eval = FALSE}
# Load the log-likelihood script
source(file = "fast_lba_ll1b.R")

# Specify the parameter vector with single threshold (b) parameter
pars <- c("A","b","vc","ve","t0")

# Specifiy a priors list
priors <- list(
  theta_mu = rep(0, length(pars)),
  theta_sig = diag(rep(1, length(pars)))
)

# Setup your sampler object - include your data, parameter vector, 
# priors and log-likelihood function
# Note: Your log-likelihood function must be loaded before this point
sampler <- pmwgs(
  data = forstmann,
  pars = pars,
  prior = priors,
  ll_func = fast_lba_ll1b
)

# Start points are not included in this example
# Initiatlise the sampler
sampler <- init(sampler)

# Burn-in stage
burned <- run_stage(sampler, 
                    stage = "burn",
                    iter = 500,
                    particles = 1000,
                    epsilon = .5)
# Adaptation stage
adapted <- run_stage(burned,
                     stage = "adapt",
                     iter = 10000,
                     particles = 1000,
                     epsilon = .5)
# Sample stage
sampled <- run_stage(adapted, 
                     stage = "sample",
                     iter = 1000,
                     particles = 200,
                     epsilon = .5)
```
<b>Note:</b> we keep the priors, start points, number of iterations and particles the same as our three threshold LBA model, so there is no bias for either model.

## Checking Descriptive Adequacy of 1b model.

The checking procedure below is the same as the three threshold LBA model, except we use the single threshold LBA model's `sampled` object/data. 

```{r part10.5.1, eval=FALSE}
load("forstmann1b_sampled.RData")
```

As we did for the three threshold LBA model, we generate 20 posterior predictive data samples. 
Note: this function is from section \@ref(genppdatafunc)
```{r part10.5.3, eval=FALSE}
pp_data_1b <- gen_pp_data(sampled, n = 20)
```

```{r part10.5.4, include=FALSE}
load("ppdata1b.RData")
```

```{r plotppdata1b, include=TRUE}
# Subject x condition Q25, median and Q75 respone time + mean accuracy
# Forstmann dataset
pq1b <- forstmann %>%
  group_by(condition, subject) %>%
  summarise(Q25 = quantile(rt, prob = 0.25),
            median = median(rt),
            Q75 = quantile(rt, prob = 0.75),
            acc =  mean(ifelse(stim == resp, 1, 0)),
            .groups = "keep"
            )

# Subject x condition Q25, median and Q75 respone time for posterior predictive data
pp_pq1b <- pp_data_1b %>%
  group_by(condition, pp_iter, subject) %>%
  summarise(Q25 = quantile(rt, prob = 0.25),
            median = median(rt),
            Q75 = quantile(rt, prob = 0.75),
            acc = mean(ifelse(stim == resp, 1, 0)),
            .groups = "keep"
            )

# Combine data with posterior predictive data and add data source
pq1b <- bind_rows(cbind(src = rep("data", nrow(pq1b)), pq1b),
                cbind(src = rep("model", nrow(pp_pq1b)), pp_pq1b)
                )

# Mean Q25, median, Q4 and accuracy for data and posterior predictive data
av_pq1b <- pq1b %>%
  group_by(src, condition) %>%
  summarise_at(vars(Q25:acc), mean)

# Variances of posterior samples
pp_var1b <- pq1b %>%
  filter(src != "data") %>%
  group_by(condition, pp_iter, src) %>%
  summarise_at(vars(Q25:acc), mean)

# Convert source column to a factor and add labels
av_pq1b$src <- factor(av_pq1b$src,
                    levels = c("model", "data"),
                    labels = c("model", "data")
                    )

pp_varsrc1b <- factor(pp_var1b$src,
                    levels = c("model", "data"),
                    labels = c("model", "data")
                    )
# Rename conditions
levels(av_pq1b$condition) <-
  levels(pp_var1b$condition) <- c("Accuracy", "Neutral", "Speed")

# Convert rt to milliseconds and acc to percentage
av_pq1b$acc <- 100 * av_pq1b$acc
pp_var1b$acc <- 100 * pp_var1b$acc
av_pq1b[, c("Q25", "median", "Q75")] <- 1000 * av_pq1b[, c("Q25", "median", "Q75")]
pp_var1b[, c("Q25", "median", "Q75")] <- 1000 * pp_var1b[, c("Q25", "median", "Q75")]
```

```{r hideppdataplots1b, echo=FALSE}
# Plot response times for data and posterior predictive data
rtplot1b <- ggplot(data = av_pq1b,
                 mapping = aes(x = condition,
                               y = Q25,
                               group = src,
                                colour = src
                               )
                 ) +
  geom_point(size = 2) +
  geom_point(mapping = aes(x = condition,
                           y = median,
                           color = src,
                           group = src
                           ),
             size = 4) +
  geom_point(mapping = aes(x = condition,
                           y = Q75,
                           color = src,
                           group = src
                           ),
             size = 2) +
  geom_point(data = pp_var1b,
             mapping = aes(x = condition,
                           y = Q25,
                           group = src),
             size = 1,
             color = "black",
             alpha = 0.05) +
  geom_point(data = pp_var1b,
             mapping = aes(x = condition,
                           y = median,
                           group = src),
             size = 2,
             color = "black",
             alpha = 0.05) +
  geom_point(data = pp_var1b,
             mapping = aes(x = condition,
                           y = Q75,
                           group = src),
             size = 1,
             color = "black",
             alpha = 0.05) +
  labs(x = "Condition",
       y = "RT Percentiles (ms)",
       colour = "Data source") +
  theme(axis.title.y = element_text(size = 14, face = 'bold'),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.75, 'cm'),
        legend.title = element_text(size = 14, face = 'bold'),
        legend.title.align = 0.5,
        legend.position = "top") +
   ylim(275, 650)

# Plot accuracy for data and posterior predictive data
accplot1b <- ggplot(data = av_pq1b,
                  mapping = aes(x = condition,
                                y = acc,
                                color = src,
                                group = src
                                )) +
  geom_point(size = 4) +
  geom_point(data = pp_var1b,
             mapping = aes(x = condition,
                           y = acc,
                           group = src),
             color = "black",
             alpha = .2
             ) +
  labs(x = "Condition",
       y = "Mean Accuracy (%)") +
  theme(axis.title.y = element_text(size = 14, face = 'bold'),
        axis.text.y = element_text(size=14),
        axis.title.x = element_text(size = 14, face = 'bold'),
        axis.text.x = element_text(size=14),
        legend.position = "none"
        ) +
  ylim(75, 90)
```

```{r ppdataplot1b, fig.align='center', fig.height=10, echo=FALSE, warning=FALSE}
plots1b <- rtplot1b/accplot1b
plots1b
```

## Model Comparison

In this section, we are going to compare the three threshold model and the single threshold model for the forstmann (2008) data to determine which model best represents the data. We begin by using a graphical method i.e. plotting the modelled data against the observed data.
 
### Assessing Descriptive Adequacy Graphically

We can assess descriptive adequacy graphically by plotting the observed data, three threshold and single threshold data on a single plot. 

```{r ppdataModelComp, include=TRUE}
# Subject x condition Q25, median and Q75 respone time + mean accuracy 
# Forstmann data
pq <- forstmann %>%
  group_by(condition, subject) %>%
  summarise(Q25 = quantile(rt, prob = 0.25),
            median = median(rt),
            Q75 = quantile(rt, prob = 0.75),
            acc =  mean(ifelse(stim == resp, 1, 0)),
            .groups = "keep"
            )

# Subject x condition Q25, median and Q75 respone time for posterior predictive data
#3b model
pp_pq3b <- pp_data_3b %>%
  group_by(condition, pp_iter, subject) %>%
  summarise(Q25 = quantile(rt, prob = 0.25),
            median = median(rt), 
            Q75 = quantile(rt, prob = 0.75),
            acc = mean(ifelse(stim == resp, 1, 0)),
            .groups = "keep"
            )
# Subject x condition Q25, median and Q75 respone time for posterior predictive data 
#1b model
pp_pq1b <- pp_data_1b %>%
  group_by(condition, pp_iter, subject) %>%
  summarise(Q25 = quantile(rt, prob = 0.25),
            median = median(rt),
            Q75 = quantile(rt, prob = 0.75),
            acc = mean(ifelse(stim == resp, 1, 0)),
            .groups = "keep"
            )

# Combine data with posterior predictive data for 1b and 3b models + add data source
pqall <- bind_rows(cbind(src = rep("data", nrow(pq)), pq),
                cbind(src = rep("3b", nrow(pp_pq3b)), pp_pq3b),
                cbind(src = rep("1b", nrow(pp_pq1b)), pp_pq1b)
                )

# Mean Q25, median, Q4 and accuracy for data and posterior predictive data
av_pq <- pqall %>%
  group_by(src, condition) %>%
  summarise_at(vars(Q25:acc), mean)

# Variances of posterior samples
pp_var <- pqall %>%
  filter(src != "data") %>%
  group_by(condition, pp_iter, src) %>%
  summarise_at(vars(Q25:acc), mean)

# Convert source column to a factor and add labels
av_pq$src <- factor(av_pq$src,
                    levels = c("3b", "data", "1b"),
                    labels = c("3b", "data", "1b")
                    )

pp_varsrc <- factor(pp_var$src,
                    levels = c("3b", "data", "1b"),
                    labels = c("3b", "data", "1b")
                    )
# Rename conditions
levels(av_pq$condition) <-
  levels(pp_var$condition) <- c("Accuracy", "Neutral", "Speed")

# Convert rt to milliseconds and acc to percentage
av_pq$acc <- 100 * av_pq$acc
pp_var$acc <- 100 * pp_var$acc
av_pq[, c("Q25", "median", "Q75")] <- 1000 * av_pq[, c("Q25", "median", "Q75")]
pp_var[, c("Q25", "median", "Q75")] <- 1000 * pp_var[, c("Q25", "median", "Q75")]
```



```{r hideppdataplotsModelComp, echo=FALSE}
# Plot response times for data and posterior predictive data
rtplot <- ggplot(data = av_pq,
                 mapping = aes(x = condition,
                               y = Q25,
                               group = src,
                                colour = src
                               )
                 ) +
  geom_point(size = 2,
             position = position_dodge(width = .5)
             ) +
  geom_point(mapping = aes(x = condition,
                           y = median,
                           color = src,
                           group = src
                           ),
             size = 4, 
             position = position_dodge(width = .5)
             ) +
  geom_point(mapping = aes(x = condition,
                           y = Q75,
                           color = src,
                           group = src
                           ),
             size = 2,
             position = position_dodge(width = .5)
             ) +
  geom_point(data = pp_var,
             mapping = aes(x = condition,
                           y = Q25,
                           group = src),
             size = 2,
             color = "black",
             alpha = 0.05,
             position = position_dodge(width = -.5)
             ) +
  geom_point(data = pp_var,
             mapping = aes(x = condition,
                           y = median,
                           group = src),
             size = 4,
             color = "black",
             alpha = 0.05, 
             position = position_dodge(width = -.5)
             ) +
  geom_point(data = pp_var,
             mapping = aes(x = condition,
                           y = Q75,
                           group = src),
             size = 2,
             color = "black",
             alpha = 0.05,
             position = position_dodge(width = -.5)
             ) +
  labs(x = "Condition",
       y = "RT Percentiles (ms)",
       colour = "Data source") +
  theme(axis.title.y = element_text(size = 14, face = 'bold'),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.75, 'cm'),
        legend.title = element_text(size = 14, face = 'bold'),
        legend.title.align = 0.5,
        legend.position = "top") +
   ylim(275, 650)

# Plot accuracy for data and posterior predictive data
accplot <- ggplot(data = av_pq,
                  mapping = aes(x = condition,
                                y = acc,
                                color = src,
                                group = src
                                )
                  ) +
  geom_point(size = 4,
             position = position_dodge(width = .5)) +
  geom_point(data = pp_var,
             mapping = aes(x = condition,
                           y = acc,
                           group = src),
             color = "black",
             alpha = 0.2,
             position = position_dodge(width = -.5)) +
  labs(x = "Condition",
       y = "Mean Accuracy (%)") +
  theme(axis.title.y = element_text(size = 14, face = 'bold'),
        axis.text.y = element_text(size=14),
        axis.title.x = element_text(size = 14, face = 'bold'),
        axis.text.x = element_text(size=14),
        legend.position = "none"
        ) +
  ylim(75, 90)
```

```{r ppdataplotModelComp, fig.align='center', fig.height=10, echo=FALSE, warning=FALSE}
ModelCompPlots <- rtplot/accplot
ModelCompPlots
```
The plots show that values in the three threshold model (red circles) better describe the data than the single threshold model (blue circles).

### Model comparison via DIC

Another method of model comparison is using information criterions. Here we provide a function for calculating deviance information criterion (DIC). <b>Note:</b> we recommend using marginal likelihood (Bayes factors) instead of DIC for model selection. For more information, see this paper on [estimating the Marginal Likelihood via importance sampling](https://link.springer.com/article/10.3758/s13428-020-01348-w).

```{r DICcode, include = TRUE}
pmwg_DIC <- function(sampled, pD = FALSE){
  # Identify number of subjects
  nsubj <- length(unique(sampled$data$subject))
 
  # Mean log-likelihood of the overall (sampled-stage) model, for each subject
  mean_ll <- apply(sampled$samples$subj_ll[, sampled$samples$stage == "sample"],
                   1,
                   mean)
 
  # Mean of each parameter across iterations. 
  # Keep dimensions for parameters and subjects
  mean_pars <- t(apply(sampled$samples$alpha, 1:2, mean))
 
  # Name 'mean_pars' so it can be used by the log_like function
  colnames(mean_pars) <- sampled$par_names
 
  # log-likelihood for each subject using their mean parameter vector
  mean_pars_ll <- numeric(ncol(mean_pars))
  
  data <- transform(sampled$data, 
                    subject = match(subject, unique(subject)))
  
  for (j in 1:nsubj) {
    mean_pars_ll[j] <- sampled$ll_func(mean_pars[j, ],
                                       data = data[data$subject == j,],
                                       sample = FALSE)
  }
 
  # Effective number of parameters
  pD <- sum(-2 * mean_ll + 2 * mean_pars_ll)
 
  # Deviance Information Criterion
  DIC <- sum(-4 * mean_ll + 2 * mean_pars_ll)
 
  if (pD){
    return(c("DIC " = DIC, " Effective parameters" = pD))
  }else{
    return(DIC)
  }
   
}
 
```

We calculate the DIC value for each model by passing the `sampled` object into the `pmwgDIC` function. 
```{r DIC3b, echo=FALSE}
load("forstmann3b_sampled.RData")
sampled3b <- sampled
```

```{r InstructDIC3b, eval=FALSE}
load("forstmann3b_sampled.RData")
sampled3b <- sampled
pmwg_DIC(sampled = sampled3b)
```

```{r print3bDIC, echo=FALSE, results = 'asis'}
pmwg_DIC(sampled = sampled3b)
```

```{r DIC1b, echo=FALSE}
load("forstmann1b_sampled.RData")
sampled1b <- sampled
```

```{r InstructDIC1b, eval=FALSE}
load("forstmann1b_sampled.RData")
sampled1b <- sampled
pmwg_DIC(sampled = sampled1b)
```

```{r print1bDIC, echo=FALSE, results='asis'}
pmwg_DIC(sampled = sampled1b)
```

The three threshold model has a lower DIC value than the single threshold model, indicating a better explanation of the data. Based on the graphical evidence and the DIC, we can be confident that the three threshold model (i.e the model where threshold is varied with speed and accuracy instructions) is a better fit to our data than the single threshold model. 

## Checking the LBA log-likelihood function

Here we show two ways to test your log-likelihood function. This method is not fail-safe, however, it allows you to establish whether your function operates as intended. 

We will use the three threshold LBA model log-likelihood function in the following examples. 


### Test one: Do changes in parameter values cause changes in the returned log-likelihood? 

We will test our log-likelihood functions by passing a parameter vector (`x`) of plausible, arbitrary values, followed by a second parameter vector with different, plausible, arbitrary values. If the log-likehood functions are functioning correctly, we should see the log-likelihood value change with the change in parameter values. We may also wish to check that unrealistic parameter values are effectively dealt with (for example, `t0` should not be able to be greater than any RTs). 

Let's compare the output from the trialwise log-likelihood function in section \@ref(writellFunc) and the fast log-likelihood function in section \@ref(fstLBALL) :
```{r llTest, include=TRUE}
x <- log(c(A = 2,
           b.1 = 2, 
           b.2 = 1, 
           b.3 = .5,
           vc = 4,
           ve = 2,
           t0 = .18
           )
         )
```

```{r, testtwll, include=TRUE}
tw_lba_ll(x, forstmann, sample = FALSE)
```

```{r, testfastll, include=TRUE}
fast_lba_ll3b(x, forstmann, sample = FALSE)
```

Great -- the log-likelihoods are the same.
Now let's change the parameter values, in particular, the three threshold parameters, and see if the log-likelihood changes. 

```{r llTestChange, include=TRUE}
x <- log(c(A = 2,
           b.1 = 1,
           b.2 = 3,
           b.3 = 3,
           vc = 4,
           ve = 2,
           t0 = .18
           )
         )
```

```{r, twTest, include= TRUE}
tw_lba_ll(x, forstmann, sample = FALSE)
```

```{r fastllTest, include=TRUE}
fast_lba_ll3b(x, forstmann, sample = FALSE)
```

These log-likelihoods are lower than the previous two, so the parameter (`x`) values are less accurate given the data. Notice how we can also use this process to check the fast, efficient log-likelihood function (`fast_lba_ll3b`) against the trialwise function (`tw_lba_ll`). Given the same data and parameter values - they should give the same output, as we see above.

<b>Not</b>: We only changed some named (`x`) parameter values. You could perform a thorough check by changing one parameter value at a time and see if there is a change in the log-likelihood. 

This is a good first check to see if our log-likelihood is behaving as it should; changing log-likelihood value with changing parameter values. However, our log-likelihood functions may contain inconsistencies, which leads us onto our second check.

### Testing whether true-x values have the highest likelihood

A more comprehensive way to test your log-likelihood function is to generate 'synthetic' data for which you know the true data generating values. This is a form of 'parameter recovery', where you run the sampler on generated data and plot the log-likelihood change as x values move away from your 'true' x-value. If the true-x has the highest log-likelihood, then it is likely that your function is working as intended. 

There are several steps to this method:

 1. Assign values to parameters of the model - these are known as the generating values. <br><b>Note</b>: ensure these are sensible values - for example in the LBA, we want the drift rate for the mismatch/error accumulator (`ve`) to be smaller than the match/correct accumulator (`vc`) and the non-decision time parameter `t0` should not be too large. 

```{r check2Par,include=TRUE}
true.x <- log(c(A = 2,
                b.1 = 2,
                b.2 = 1,
                b.3 = .5,
                vc = 4,
                ve = 2,
                t0 = .18
                )
              )
```

 2.  Pass the generating parameter values (`true.x`) into the log-likelihood and generate data using your log-likelihood function, where `sample = TRUE`.
```{r testdatacheck2, include=TRUE}
test.data <- tw_lba_ll(x = true.x,
                       data = data,
                       sample = TRUE
                       )
```

 3. Generate profile plots to compare the log-likelihood of the `true.x` parameter values against nearby values in the 'synethetic' data. You can write your own function for this, or use the function below. 

Our plotting function takes five arguments: 

* `x` is a vector of named parameter values.
* `n.values` specifies how many parameter x-values we will assess to compare to compare to the true data-generating parameter value. Default is 9.
* `synth.data` is the synthetic data
* `dims` specifies the layout for the plots we will generate. In this example, we have 7 parameters, so I will set my dimensions to be 4 rows with 2 columns.
* `ll_func` specifies the name of the log-likelihood function you have created.
* `server` should only be set to true if running on a device with multiple cores. This argument speeds up computational efficiency.
 

```{r testllfun,include=TRUE}

library(parallel)

test.likelihood <- function(x, 
                            n.values = 9,
                            synth.data,
                            dims,
                            ll_func,
                            server = FALSE,...) {
  pars <- names(x)
  x.tmp <- x
  sequence <- seq(from = -.22, to = .22,
                  length.out = n.values)
  op <- par(mfrow = dims)
  par_likelihoods <- lapply(setNames(pars, pars), function(p){
    testvalues <- sequence + true.x[[p]]
    tmp <- unlist(lapply(testvalues, function (i){ 
      #for each test vlaue, it will apply the function where i = testvalue.
    x.tmp[[p]] <- i
    ll_func(x = x.tmp, data = synth.data)
    }))
    
    plot(x = testvalues,
         y = tmp,
         type = "b",
         main = p,
         xlab = "log par values",
         ylab = "log-likelihood",
         ...)
    abline(v = true.x[[p]], col = "red")
    return(tmp)})
  par(op)
  return(par_likelihoods)
}
```

Whilst we used the trialwise log-likelihood function to generate data, we will use the fast log-likelihood function in this "parameter recovery" simulation for computational efficiency. 

```{r llPlots, include=TRUE}
ll_plots <- test.likelihood(x = true.x,
                                    n.values = 9,
                                    synth.data = test.data,
                                    dims = c(2, 2),
                                    ll_func = fast_lba_ll3b
                                    ) 
```

The red line indicates the true generating parameter value (i.e. the data generating values for the parameters) have the best (highest) likelihoods.

Note: The plots above do not tell you if your log-likelihood is correct in the sense that you have written a log-likelihood for the model you actually want to test. The plots allow you to determine if your function is functioning as intended.

<b> Explanation needs to be improved - more detail required. the second sentence seems tacked on.</b>
Below is an example output from a function where there is a potential error. Another potential error is where the lines are flat. 

```{r dodgyll, include=FALSE}

incorrect_lba_ll <- function(x, data, sample = FALSE) {
  x <- exp(x)
  if (any(data$rt < x["t0"])) {
    return(-1e10)
  }
  
  if (sample) {
    data$rt <- rep(NA, nrow(data))
    data$resp <- rep(NA, nrow(data))
  } else {
    out <- numeric(nrow(data))
  }
  
  if (sample) { 
  for (i in 1:nrow(data)) {
    A = x["A"]
    b = x[paste0("b.", data$condition[i])] + A
    vc = x["vc"]
    ve = x["ve"]
    t0 = x["t0"]
    s = c(1, 1)
    if (data$stim[i] == 1) {
      vs = c(vc, ve)
    } else {
      vs = c(ve, vc)
    }
  
      tmp <-
        rLBA(
          n = 1,
          A = A,
          b = b,
          mean_v = vs,
          sd_v = s,
          t0 = t0,
          dist = "norm",
          silent = TRUE
        )
      data$rt[i] <- tmp$rt
      data$resp[i] <- tmp$resp
    } 
  } else {
    
  all_b = numeric(nrow(data))
  vlist = list("v.1" = numeric(nrow(data)), "v.2" = numeric(nrow(data)))
  
  
  stim = levels(data$stim)
  con = levels(data$condition)
  
  for (c in con) {
    for (s in stim) {
      use <- data$condition == c &
        data$stim == s
      if (any(use)) {
        bs = x[paste0("b.", c)] + x["A"]
        all_b[use] = bs
        vc = x["vc"]
        ve = x["ve"]
        if (s == 1) {
          vlist$v.1[use] = ve
          vlist$v.2[use] = vc
        } else {
          vlist$v.1[use] = vc
          vlist$v.2[use] = ve
        }
      }
    }
  }
    
   out <- dLBA(
      rt = data$rt,
      response = data$resp,
      A = x["A"],
      b =  all_b,
      mean_v = vlist,
      sd_v = c(1, 1),
      t0 = x["t0"],
      distribution = "norm",
      silent = TRUE
    )
  
  }
  
  if (sample) {
    return(data)
  } else {
    bad <- (out < 1e-10) | (!is.finite(out))
    out[bad] <- 1e-10
    out <- sum(log(out))
    return(out)
  }
}

```

```{r dodgellplots, echo=FALSE}
incorrect.ll.plots <- test.likelihood(x = true.x,
                                      n.values = 9,
                                      synth.data = test.data,
                                      dims = c(2,2),
                                      ll_func = incorrect_lba_ll
                                      ) 
```

