# PMwG sampler and sequential sampling models


In this chapter we'll demonstrate how to use the PMwG sampler with a sequential sampling model; the Linear Ballistic Accumulator (LBA).
Please ensure the `PMwG` and `rtdists` packages are installed. We currently recommended installing `PMwG` via devtools.
```{r getpkg, eval=FALSE}
# The pmwg package will be on CRAN - this step will be removed.
install_github('newcastlecl\pmwg')
```
```{r loadpkg}
library(pmwg)
```

## Description of Forstmann experiment

Forstmann et al looked at neural correlates of decision making under time pressure, with an aim to identify areas of the brain associated with speed-accuracy tradeoff. Imaging (fMRI) and behavioural data was collected; however, we will analyse behavioural data from the decision-making task only. In terms of modelling the data, Forstmann expected to find differences in thresholds (direction?) for each of the three speed-emphasis conditions. We have included the Forstmann et als data in the `PMwG` package as a data frame named `forstmann`. The sampler requires a data frame with a <b>`subject`</b> column. The subject column data type can be a factor or numeric.

Table \@ref(tab:forsthead10) shows the first ten trials from the Forstmann dataset. Participants `(n = 19)` were asked to indicate whether a cloud of dots in a random-dot kinematogram (RDK) moved to the left or the right of the screen. The IV was a within-subject, speed-accuracy manipulation where, before each trial began, pariticipants were instructed to make their choice <i>accurately</i> `(condition = 1)`, with <i>urgency</i>`(condition = 3)`or were presented with a <i>neutral</i> message `(condition = 2)`. Stimuli moved either <i>left</i> `(stim = 1)` or <i>right</i> `(stim = 2)` and responses were <i>left</i> `(resp = 1)` or <i>right</i> `(resp = 2)`. Response times `(rt)` were recorded in seconds. For more information about the design of the experiment please see [the original paper](https://www.pnas.org/content/105/45/17538.short).

```{r forsthead10, echo=FALSE, out.width='80%', fig.asp=.75, fig.align='center'}
kable(head(forstmann, 10), row.names = F,  align = "ccccc", caption = 'First 10 trials in Forstmann dataset. The `forstmann` dataset is an object/data frame ')
```

## Linear Ballistic Accumulator Parameters

There are preliminary steps we need to complete before running the sampler. Let's begin by defining the Linear Ballistic Accumulator (LBA) [@brown2008simplest] model parameters.

* `b` threshold parameter (the evidence required to trigger a response)
* `v` is drift rate or average speed of evidence accumulation
* `A` is the model's start point 
* `t0` is non-decision time 
* `sv` is the standard deviation of drift rate

## Writing the log-likelihood function

Just as we did with the SDT example, we'll write a slow and a fast log-likelihood function. The runtime difference is caused by calling the `dLBA` function line-by-line for the slow log-likelihood and calling the `dLBA` function once for all the data in the fast log-likelihood function. When writing a new log-likelihood function, we suggest starting with a slow, line-by-line function for easier debugging. 


The LBA log-likelihood function takes three arguments: 

* `x` is a named parameter vector (e.g. `pars`)
* `data` is your dataset (e.g.`forstmann`). Your dataset must include a `"subject"` column
* `sample = FALSE` calculates a density function or `TRUE` generates a posterior, predictive sample that matches the shape of data.


The log-likelihood function shown below includes functions from the `rtdists` package for generating data and estimating density. If you'd like to run through this example, it is best to copy the `tw_lba_ll` function from the code block below rather than copying from the separate code chunks where curly braces have been removed.

NOTE: The trialwise log-likelihood is very slow and inneficient because `rLBA` and `dLBA` will be called on each line of the data. This will result in very slow sampling times and is a consequence of the `rtdists` package, not an issue with the PMwG sampling speed. If you have experience writing log-likelihoods, we recommend writing a faster version than our trialwise function, or use the fast log-likelihood we have written in section \@ref(fstLBALL). If you are new to modelling, we recommend trying the trialwise (slow) log-likelihood function as it is easier to follow, troubleshoot and is less likely to result in errors. 

Let's begin by installing `rtdists` package...

```{r instrtdist}
library(rtdists)
```

and now our complete trialwise (slow) log-likelihood function.

```{r trialwiseLL, attr.source = '.numberLines', include=TRUE }
tw_lba_ll <- function(x, data, sample = FALSE) {
  x <- exp(x)
  if (any(data$rt < x["t0"])) {
    return(-1e10)
  }
  
  if (sample) {
    tmp <- numeric(nrow(data))
    data$rt <- rep(NA, nrow(data))
    data$resp <- rep(NA, nrow(data))
  } else {
    out <- numeric(nrow(data))
  }
  
  for (i in 1:nrow(data)) {
    A = x["A"]
    b = x[paste0("b.", data$condition[i])] + A
    vc = x["vc"]
    ve = x["ve"]
    t0 = x["t0"]
    s = c(1, 1)
    
    if (data$stim[i] == 1) {
      vs = c(vc, ve)
      } else {
      vs = c(ve, vc)
      }
    
    if (sample) {
      tmp <- rLBA(n = 1,
                  A = A,
                  b = b,
                  mean_v = vs,
                  sd_v = s,
                  t0 = t0,
                  dist = "norm",
                  silent = TRUE
                  )
      data$rt[i] <- tmp$rt
      data$resp[i] <- tmp$resp
    } else {
      out[i] <- dLBA(rt = data$rt[i],
                     response = data$resp[i],
                     A = A,
                     b = b,
                     mean_v = vs,
                     sd_v = s,
                     t0 = t0,
                     dist = "norm",
                     silent = TRUE
      )
      }
  }
  
if (sample) {
    return(data)
  } else {
    bad <- (out < 1e-10) | (!is.finite(out))
    out[bad] <- 1e-10
    out <- sum(log(out))
    return(out)
  }
}
```


The first line in the `tw_lba_ll` function (Line 2 below) takes the exponent of the parameter values. We do this as the LBA expects values on the real number line. Line 3 and 4 then checks RTs are faster than the non-decision time parameter `t0`, and return a low value, indicating that the given value of `t0` is not likely. 

```{r twLBAll1, attr.source = '.numberLines startFrom="2"', echo=TRUE, eval=FALSE}
  x <- exp(x)
  if (any(data$rt < x["t0"])) {
    return(-1e10)
```


Now we create a vector with values sampled from the posterior distribution OR estimating the density. If `sample = TRUE`, we remove all responses `(resp)` and rts.  This means when we return `data`, we are returning the posterior predictive data which matches with the associated `subject` and `condition`.

If `sample = FALSE` (the `else` statement from line 11) we create an `out` vector, with a length equal to the number of rows in the dataset, and store the likelihood value for each subject and condition.

```{r twLBAll2, attr.source = '.numberLines startFrom="7"', echo=TRUE, eval=FALSE}
  if (sample) {
    tmp <- numeric(nrow(data))
    data$rt <- rep(NA, nrow(data))
    data$resp <- rep(NA, nrow(data))
  } else {
    out <- numeric(nrow(data))
  }
```


Next, we loop over rows in the dataset. In this `for` loop, we find the values `(x)` of each parameter in our model for each row, so that any conditional parameters (for example `b` in our model) are correctly assigned. For example, we want to calculate the density for a model that has three threshold parameters (one for each condition; `1  = accuracy`, `2 = neutral`, or `3 = speed`). In the loop, we paste `b.` to the condition in row `[i]` and add `A` (the start point - we do this to ensure the threshold is greater than the starting point). 

On line 22 we set the order of our drift rate parameters. Recall that `stim = 1` is a stimulus moving to the left. `dLBA` requires the drift accumulators to be matched i.e. when `data$stim[i] == 1`, the drift rate for the correct accumulator `(vc)` is in position one, so we order the drift rates; `vs = c(vc, ve)`. The `else` statement addresses right moving stimuli `data$stim[i] == 2`, the incorrect accumulator `(ve)` is the first accumulator, so the drift rate parameter order is `vs = c(ve, vc)`. This ensures that the correct `(vc)` and error `(ve)` drift rates match with the corresponding accumulators for given stimuli. 

```{r twLBAll3, attr.source = '.numberLines startFrom="15"', echo=TRUE, eval=FALSE}
  for (i in 1:nrow(data)) {
    A = x["A"]
    b = x[paste0("b.", data$condition[i])] + A
    vc = x["vc"]
    ve = x["ve"]
    t0 = x["t0"]
    s = c(1, 1)
    if (data$stim[i] == 1) {
      vs = c(vc, ve)
    } else {
      vs = c(ve, vc)
    }
```

The following section calls the relevant `rtdists` function depending on whether we are sampling from the posterior `(rLBA)` or estimating the density `(dLBA)`. We then input the parameters from above (using the names set above) into the relevant function. When generating data from the posterior (Line 30-37), `rLBA` is called for each line of the data, storing the generated `rt` and `response` given the posterior parameter estimates in the `tmp` vector (which we then reassign to the empty `data$rt` and `data$resp` columns). We set `n = 1`, since we are calling `rLBA` on 1 row of the data. When estimating the density (Line 42 to 50), dLBA is called for each line of the data, storing the probability of the `rt` and `response` under the proposed parameters `(x)` in the `out` vector. 


```{r lbaLL3, attr.source='.numberLines startFrom="29"', echo=TRUE, eval=FALSE}
  if (sample) {
      tmp <- rLBA(n = 1,
                  A = A,
                  b = b,
                  mean_v = vs,
                  sd_v = s,
                  t0 = t0,
                  dist = "norm",
                  silent = TRUE
                  )
      data$rt[i] <- tmp$rt
      data$resp[i] <- tmp$resp
    } else {
      out[i] <- dLBA(rt = data$rt[i],
                     response = data$resp[i],
                     A = A,
                     b = b,
                     mean_v = vs,
                     sd_v = s,
                     t0 = t0,
                     dist = "norm",
                     silent = TRUE
      )
      }
```

This final section tells the function what to return; `data` - when sampling posterior predictive (Line 56) - or the sum of the likelihoods - when estimating density (Line x - y). On line 58 we take all implausible likelihood values, assign them to the `bad` object and then (line 59) set them to zero. The final two lines within the `else` statement take the log of all likelihood values, sums them, assigns the model’s log-likelihood value to the `out` variable and returns that value.

```{r lbaLL43, attr.source='.numberLines startFrom="55"', echo=TRUE, eval=FALSE}
if (sample) {
    return(data)
  } else {
    bad <- (out < 1e-10) | (!is.finite(out))
    out[bad] <- 1e-10
    out <- sum(log(out))
    return(out)
  }
```

### Fast LBA Log-likelihood Function {#fstLBALL}

As the data is large, and the `dLBA` function takes some time to run, the log-likelihood code above is computationally inefficient. There are several ways to improve the log-likelihood's performance; in our example below, we reduce the number of calls to `dLBA` (and `rLBA` when sampling) to one call. We do this by passing a list of `dLBA` parameter values for the length of the data. 

<b>NOTE:</b> When generating posterior predictive data, the `rLBA` function is still executed for each row of data; however,
it is only executed several times, so computational efficiency is uncompromised.


```{r fast_lba_ll, include=TRUE}
fast_lba_ll <- function(x, data, sample = FALSE) {
  x <- exp(x)
  if (any(data$rt < x["t0"])) {
    return(-1e10)
  }
  if (sample) {
    data$rt <- rep(NA, nrow(data))
    data$resp <- rep(NA, nrow(data))
  } else {
    out <- numeric(nrow(data))
  }
  if (sample) {
    for (i in 1:nrow(data)) {
      A = x["A"]
      b = x[paste0("b.", data$condition[i])] + A
      vc = x["vc"]
      ve = x["ve"]
      t0 = x["t0"]
      s = c(1, 1)
      if (data$stim[i] == 1) {
        vs = c(vc, ve)
      } else {
        vs = c(ve, vc)
      }
      
      tmp <- rLBA(n = 1,
                  A = A,
                  b = b,
                  mean_v = vs,
                  sd_v = s,
                  t0 = t0,
                  dist = "norm",
                  silent = TRUE
                  )
      data$rt[i] <- tmp$rt
      data$resp[i] <- tmp$resp
    }
  } else {
    all_b <- numeric(nrow(data))
    vlist <- list("v.1" = numeric(nrow(data)), 
                 "v.2" = numeric(nrow(data)))
    stim <- levels(data$stim)
    con <- levels(data$condition)
    
    for (c in con) {
      for (s in stim) {
        use <- data$condition == c & data$stim == s
        if (any(use)) {
          bs = x[paste0("b.", c)] + x["A"]
          all_b[use] = bs
          vc = x["vc"]
          ve = x["ve"]
          if (s == 1) {
            vlist$v.1[use] = vc
            vlist$v.2[use] = ve
          } else {
            vlist$v.1[use] = ve
            vlist$v.2[use] = vc
          }
        }
      }
    }
    
    out <- dLBA(rt = data$rt,
                response = data$resp,
                A = x["A"],
                b =  all_b,
                mean_v = vlist,
                sd_v = c(1, 1),
                t0 = x["t0"],
                distribution = "norm",
                silent = TRUE
    )
    }
  
  if (sample) {
    return(data)
  } else {
    bad <- (out < 1e-10) | (!is.finite(out))
    out[bad] <- 1e-10
    out <- sum(log(out))
    return(out)
  }
}
```

You should improve your log-likelihood's performance as you see fit.  When you're confident that your log-likelihood code functions correctly, we suggest saving it as a separate script so it can be sourced and loaded when running the sampler. If you're learning how to write log-likelihood functions, take a look at our [troubleshooting section](#troubleshoot) for tips. 

## PMwG Framework

Now that we have a log-likelihood function, we can set up the `PMwG` sampler. Running the sampler follows the same procedure outlined in the [SDT chapter](#sdtPMwG); we need to set up a vector of model parameter names, create a `priors` object, source our LBA log-likelihood script and then create our `sampler` object.


Let's begin by creating a vector of model parameter names, which we'll use in our log-likelihood function. You can name this object as you wish; however, in our example, we name it `pars`. <br><b>IMPORTANT NOTE:</B> The parameters you list in the `pars` vector <b>must</b> match those included in your log-likelihood function i.e. they must have the same names and the same number of parameters.

For the `forstmann` dataset, we use three threshold parameters (one `b` for each condition) because we assume that the condition has an effect on level of caution. We include two drift rate parameters: `v1` for the incorrect accumulator and `v2` for the correct accumulator, a start point parameter `A` and a non-decision time `t0` parameter. 

We've made a decision to set the `sv` to 1 to satisfy the scaling properties of the model. As such, we haven't included the `sv` parameter in the `pars` vector - it is found in the LBA's log-likelihood function (see above).

```{r pars}
pars <- c("b1", "b2", "b3", "A", "v1", "v2", "t0")
```


Next we create our `priors` object; a list that contains two components:

* `theta_mu_mean` a vector containing the prior for model parameter means
* `theta_mu_var` the prior covariance matrix for model parameters.

<b>The `theta_mu_mean` object is initiated with zeros, and the covariance  has the diagonal of 1 and 0's for the off diagonal. We use generic priors here which are broad. This means that the prior is <b>blah</b> and the covariance matrix posterior space is able to be explored. In some examples, the prior will need to be adjusted to account for different factors or priors we already know. </b> 


```{r priors}
priors <- list(theta_mu_mean = rep(0, length(pars)),
  theat_mu_var = diag(rep(1, length(pars)))
)
```

Now source and load your log-likelihood script <i>before</i> you create the sampler object.
```{r loadLL, echo=TRUE, eval=FALSE}
source(file = "fast_lba_ll.R")
```

Next we specify the PMwG `sampler` object. 
The `pmwgs` function takes a set of arguments (listed below) and returns a list containing the required components for performing the particle metropolis within Gibbs steps.

* `data =` your data - a data frame (e.g.`forstmann`) with a column for participants called <b>`subject`</b>
* `pars =` the model parameters to be used (e.g.`pars`)
* `prior =` the priors to be used (e.g.`priors`)
* `ll_func =` name of log-likelihood function to be used (e.g.`fast_lba_ll`)

```{r forstSampObj, eval=FALSE}
sampler <- pmwgs(data = data,
                 pars = pars,
                 prior = priors,
                 ll_func = fast_lba_ll
                 )
```


### Model start points {#lbaStartPts}
There is also an option to set model start points. We have specified sensible start points for the `forstmann` dataset under the LBA model. If you choose not to specify start points, the sampler will randomly sample points from the prior distribution. 

The `start_points` object contains two vectors:

* `mu` a vector of start points for the mu of each model parameter 
* `sig2` vector containing the start points of the covariance matrix of covariance between model parameters.

NOTE: Start points must be on the real line. Our log-likelihood function immediately takes the exponent of these start points and returns only positive values, so we use the log of sensible start points here. 

```{r lbaStartPts, echo=TRUE, eval=FALSE}
start_points <- list(
  mu = c(log(1.2), log(1.2), log(1.2), log(1.4), log(1.3), log(3.5), log(0.13)),
  sig2 = diag(rep(.01, length(pars)))
)
```


### Running the sampler {#run-sampler}

The sampler setup is now complete and we can run the sampler. First we use the `init` function to generate initial start points for the random effects and store them in the `sampler` object. Here we specify start points by providing values for the `start_mu` and `start_sig` arguments.

```{r runlbaSampler, echo=TRUE, eval=FALSE}
sampler <- init(sampler, start_mu = start_points$mu,
                start_sig = start_points$sig2)
```


To run the sampler, we use the `run_stage` function. The `run_stage` function takes four arguments:

* `x =` the `sampler` object including parameters that was created from the `init` function above.
* `stage =` the sampling stage (These are:`"burn"`, `"adapt"` or `"sample"` - in this order).
* `iter =` is the number of iterations for the sampling stage. This is similar to running deMCMC, where it takes many iterations to reach the posterior space. Default = 1000.
* `particles =` is the number of particles generated on each iteration. Default = 1000.
* `epsilon =` is a value between 0 and 1 which reduces the size of the sampling space. We use lower values of epsilon when there are more parameters to estimate. Default = 1.
* `n_cores =` the number of cores on a machine you wish to use to run the sammpler. This allows sampling to be run across cores (parallelising for subjects). Default = 1. 

The speed of the sampling process is dependent upon the number of cores used i.e. the more cores you use, the faster sampling proceeds.


<!-- This needs to be added to troubleshoot page with references from each other page

NOTE: We advocate running burn-in for a low number of iterations and particles first. This will give you a sense of a) whether the sampler is working as intended (see troubleshooting/checks for what parameter chains should look like), b) the number of iterations & particles needed to achieve the target acceptance rate, as well as the appropriate epsilon value. The acceptance rate is generally very high for the first few iterations ( > 100) and then declines. By doing a short run, you can check and then optimise the number of particles being used (and the balance with epsilon), so that acceptance rate is close to 30% on a longer, full run. We recommend you start with epsilon = .5 to increase efficiency, then adjust as needed.  -->

<!-- NOTE: Overall, we aim for ~30% acceptance rate of particles. Acceptance rates which are too high may be inaccurate if not run for long enough. Acceptance rates which are too low will be inefficient and may fail to create an efficient distribution for the sampling stage.  -->


<!-- To run burn in, set the stage argument to "burn". -->



