# First (simple) example - Signal Detection Theory

Here we run through a simple signal detection example with data from @wagenmakers2008diffusion. For more information about the design of the experiment please see [the original paper](https://www.sciencedirect.com/science/article/pii/S0749596X07000496).

While it is unnecessary to use the sampler package for this SDT example, we hope that this demonstrates the usefullness of the samplers package.  

```{r, eval=TRUE, include=FALSE}
library("tidyverse")
library("reshape2")
# Samplers project - SDT example
parEst <- c(C.w = 0.9269677, C.nw = 1.7675702, HF.d = 2.9327993, LF.d = 2.6904818, VLF.d = 2.1823751)
# estimate curve for non-words
nwD <- round(rnorm(n = 5000, mean = 0,sd = 1), 2)
# estimate curve for high frequency words
hfwD <- round(rnorm(n = 5000, mean = parEst[3], sd = 1), 2)
# estimate curve for low frequency words
lfwD <- round(rnorm(n = 5000, mean = parEst[4], sd = 1), 2)
# estimate curve for very low frequency words
vlfwD <- round(rnorm(n = 5000, mean = parEst[5], sd = 1), 2)
# combine into data frame
dfff <- bind_cols(hf = hfwD, lf = lfwD, nw = nwD, vlf = vlfwD)
dfff <- melt(dfff)
```
```{r, echo=FALSE}
# plot distributions with non-word criterion (red line) and word criterion (blue line)
ggplot(dfff,aes(x=value, fill=variable)) + 
  geom_density(alpha=0.25) + 
  geom_vline(xintercept = parEst[1], colour = 'blue') + 
  geom_vline(xintercept = parEst[2], colour = 'red') + 
  scale_y_continuous(limits = c(0, 0.42), expand = expand_scale(mult = c(0, .1)))
```

<b>A second example might cover another SDT example with the addition of trial level covariate i.e. analytic solution NA.</b>
