# How to use the sampler - Signal Detection Theory

Here we demonstrate how to use the sampler package to run a simple signal detection theory analysis on a lexical decision task taken from  @wagenmakers2008diffusion. For details about the design of the experiment please see [the original paper](https://www.sciencedirect.com/science/article/pii/S0749596X07000496). <b> Is this the correct paper to reference? </b>

While it is unnecessary to use the sampler package for a simple analysis such as this, we hope this example demonstrates the usefullness of the samplers package.  

First we need to install the psamplers package. We currently recommended installing psamplers via devtools.
```{r getSampPkg, eval=FALSE}
# The samplers package will be on CRAN - this step will be removed.
install_github('gjcooper\samplers')
```
```{r loadSampPkg}
library(psamplers)
```

<b>We then need to get the Wagenmakers data from somewhere? Will it be included in the samplers package?</b>

```{r, eval=TRUE, include=FALSE}
library("tidyverse")
library("reshape2")
# Samplers project - SDT example
parEst <- c(C.w = 0.9269677, C.nw = 1.7675702, HF.d = 2.9327993, LF.d = 2.6904818, VLF.d = 2.1823751)
# estimate curve for non-words
nwD <- round(rnorm(n = 5000, mean = 0,sd = 1), 2)
# estimate curve for high frequency words
hfwD <- round(rnorm(n = 5000, mean = parEst[3], sd = 1), 2)
# estimate curve for low frequency words
lfwD <- round(rnorm(n = 5000, mean = parEst[4], sd = 1), 2)
# estimate curve for very low frequency words
vlfwD <- round(rnorm(n = 5000, mean = parEst[5], sd = 1), 2)
# combine into data frame
dfff <- bind_cols(hf = hfwD, lf = lfwD, nw = nwD, vlf = vlfwD)
dfff <- melt(dfff)
```
```{r, echo=FALSE}
# plot distributions with non-word criterion (red line) and word criterion (blue line)
ggplot(dfff,aes(x=value, fill=variable)) + 
  geom_density(alpha=0.25) + 
  geom_vline(xintercept = parEst[1], colour = 'blue') + 
  geom_vline(xintercept = parEst[2], colour = 'red') + 
  scale_y_continuous(limits = c(0, 0.42), expand = expand_scale(mult = c(0, .1)))
```

<b>A second example might cover another SDT example with the addition of trial level covariate i.e. analytic solution NA.</b>
